% !TEX root = ../thesis.tex
\chapter{Background}

\section{Analyzing Algorithms}

An \textit{algorithm} is a list of specific instructions written to achieve some
goal. The word ``algorithm" has become an increasingly prevalent phrase on the
internet with the popularity of social media, and it is often framed as an
opaque force of technology that controls what content users are shown.
While those social media algorithms exist, they are only one type of
algorithm. A cookie recipe is an algorithm, providing instructions a baker
follows to yield a batch of cookies upon completion. GPS devices provide
algorithms for driving from one location to another. In computer science, an
algorithm is often implemented as a sequence of lines of code that perform a computation, usually
mathematical calculations or processing data. Because algorithms are so
versatile, and there are so many different ways of accomplishing one goal, we
analyze algorithms by their efficiency: how many resources they use and how much
time they take to complete the task.

\subsection{Big-O}

To analyze efficiency, we have to define what we mean for an algorithm to be
“efficient.” For the sake of this thesis, we will focus on the time aspect of
efficiency. The amount of time an algorithm takes to execute from start to
finish is known as the \textit{running time} of an algorithm. We can
mathematically formalize running time with what’s known as \textit{Big-O}
notation.

Big-O notation is one form of \textit{asymptotic} analysis, or analysis that takes
place as the size of the input becomes infinitely large. If we want to know how
many resources an algorithm will require (resources can mean power, time,
storage space, etc.), we want to know how it performs in the \textit{worst
case} scenario, as any other case will use fewer resources. In many scenarios,
the worst case is abstractly thought to be an infinitely large input,
representing how a realistically large input often requires more computational
resources. So, worst case analysis as a facet of asymptotic analysis allows us
to guarantee an algorithm scales well for arbitrarily large inputs. 
For instance, if I am following an
algorithm to bake cookies and the steps involve rolling out the dough, mixing,
and measuring ingredients by hand, the algorithm works well for me to make a
batch of 24 cookies. However, if I start a popular cookie franchise and suddenly
I need 2,400 cookies a day, my original algorithm is not going to be the most
efficient way to make that many cookies. While it may make sense when baking
cookies to handle these two cases differently, in computer science problems
data commonly scales to very large sizes, so it is helpful to know our
algorithms will still be efficient in those cases.

We \textit{estimate} running time through worst case asymptotic analysis. In Big-O
notation, we write that for some algorithm called $f$, $f(n) = O(g(n))$. This is
read “$f(n)$ is big $O$ of $g(n)$,” where $f$ on input $n$ will take \textbf{at
most} $c \cdot g(n)$ time to run for some constant $c$. The $O$ signifies
the “at most” in that statement, or more formally it denotes an \textit{upper
bound} on the asymptotic running time. A ``constant" in this case is simply a
positive number that does not change as $n$ changes, and in Big-O we denote
$f(n) = O(c \cdot g(n)) = O(g(n))$ to indicate that we are only concerned with how
$f$ grows as $n$ becomes large. We can think of $c$ in the context of
running an algorithm on actual computers: a computer from 1990 (when
microprocessors were only just becoming accessible to the public) will have much
less processing power than a computer from 2020 with 30 more years of microprocessor
innovation. So, $c$ will be larger for the 1990 computer than the 2020 computer.
However, both computers will take more time as $n$ grows large regardless of
$c$, which is why we focus on $n$ in Big-O. It is also relevant to note that
``time" is purposefully abstract in the term ``running time;" we just use it to mean
``some unit of time." We can measure units in seconds, milliseconds, operations,
etc., so $c$ also represents the fact that we are not always using the same unit
of measurement when performing $f$. However, similarly to the processing power
of different computers, we will use more of our unit of measurement for any unit
as $n$ grows large. For example, if it takes one minute to make one batch of
cookies, it will take 10 minutes to make 10 sequential batches of cookies.
Alternatively, if it takes one hour to make a batch of cookies, it will take 10
hours to make 10 sequential batches of cookies. In both cases, if we are
committed to making cookies, the amount of time the cookies take
will grow with the number of batches ($n$) we want to perform. 

Running time can be classified into different categories. The most
relevant categories to this thesis are constant time, polynomial time, and
exponential time. In very, very simplified terms, for a basic understanding of
these categories, we can consider constant time to be fast, polynomial time to
be medium speed, and exponential time to be slow. An algorithm can belong to one
these different categories based on its worst case running time. 

Constant time, denoted $O(1)$, can be abstractly thought of as any
operation that takes the same fixed amount of time regardless of the input size.
Constant time is the fastest category of running time. Say our friend Alice
chooses to eat at Excellent Cuisine, her favorite restaurant, every day regardless of how many choices
she is presented with. If we give her a list of 10 versus 100 restaurants, she
won't even look at them, and will always choose to eat at Excellent Cuisine. We
could say Alice is following a constant time algorithm, where it always takes
her the same amount of time to decide on a restaurant no matter the length of
the list of restaurants we provide her with. Most algorithms
involve more complicated operations than those that can be done in constant
time, however, so our next category of time analysis is polynomial time.

Say Bob is also choosing a restaurant, but he likes to research every restaurant
on the list for 3 minutes before making a final decision. If we give him a list
of 10 restaurants, it will take him 30 minutes to choose. If we give him a list
of 100 restaurants, it will take him 300 minutes to choose. For any list of
restaurants of length $n$, we know it will take him $3n$ minutes to choose.
Recall that a polynomial is a mathematical expression that combines terms of
variables and constants using various operations (e.g. $x^4 + 2x^3 + 7$ or $x^2
+ y^2$). In Bob's case, if we call his algorithm $b$, then we can say $b(n) =
O(3n) = O(n)$ to indicate that the amount of time Bob takes to choose a restaurant
scales with the length $n$ of the list of restaurants. If Bob took 3 hours per
restaurant instead of 3 minutes, we would still say $b(n) = O(3n) = O(n)$. Another way to
think about polynomial time, is to consider how many
actions must be completed per input to the algorithm. For instance, if I already
have cookie dough and I need to bake the cookies, my algorithm might be: for
every tablespoon of cookie dough, roll the dough into a ball and then sprinkle
cinnamon on top. That is 2 actions to complete for every tablespoon of cookie
dough. So, if we have $n$ tablespoons, we could say this step takes $O(2n) =
O(n)$ time. If I then compared each ball of dough to every other ball to make sure
they were all the same size, it would mean that for each of the $n$ balls I
have to check it against 
$n - 1$ other balls. In this case, I am performing $n^2$ additional actions, which
we can describe as a polynomial number of actions (which is why we say it takes
polynomial time). So, I can say this whole process takes $O(n^2 + 2n) = O(n^2)$.
Because $n^2$ grows much faster than $n$, we can drop the $2n$ term in Big-O notation
to indicate $n^2$ is the relevant upper bound. 
For our purposes, we can consider polynomial
time to be sufficiently “efficient,” with the consideration that we would hope
for our algorithm to be much closer to $O(n)$ time than $O(n^{100})$, as the
former is much more efficient than the latter.

Our last category is exponential
time, which is very slow compared to our other two categories. We consider an
algorithm “inefficient” if it takes exponential time. An example of exponential
time in Big-O is $O(2^n)$. As $n$ grows larger in this case, the running time
increases extremely rapidly. To compare polynomial time and exponential time,
imagine we have some number $n = 1,000$. If we square $n$, we get a polynomial
expression: $n^2 = 1,000^2 = 1,000,000$. That is indeed large, but not compared to
$2^n = 2^{1,000}$,
which is “a number much larger than the number of atoms in the universe”
\cite{Sipser_2006}.

\subsection{P vs NP}

In time complexity theory, the field of mathematical analysis centered on the
running time of algorithms, we give more specific names to the aforementioned categories.
The most important complexity category for our purposes is $P$,
which denotes the class of problems that can be solved with polynomial time
algorithms. In other words, we can think of $P$ as a collection of problems where every
problem is solved by an algorithm that runs in polynomial time. Perhaps
intuitively, we know there exist many problems
that take longer than polynomial time to solve (such as those that take
exponential time and therefore belong to the class $Exp$). A less intuitive fact is that there
also exist problems that we think do not belong in $P$, but we don't know how to
prove it at this time nor do we believe we will know how to prove it anytime
soon.

To understand what I mean by this, we need to discuss another important
complexity class called $NP$: the class of problems solved by
\textit{nondeterministic polynomial} time algorithms. In very simplified terms,
one could think of “nondeterministic” as a brute force approach to difficult
problems: for a problem with $2^n$ possible solutions, we go through all $2^n$ of them until we find a
correct solution (assuming a correct solution exists). In the worst case, this
approach will take at least $2^n$ actions, which we now know is not polynomial time and
therefore not a feasible way to solve the problem. However, we can also think of
nondeterminism to mean that if we were told which of the $2^n$ solutions is the
correct one, we could could \textit{verify} that it is correct in polynomial
time. As an example, lets think about the classic 52-card game of solitaire \cite{solitaire}. 

\begin{figure} [H]
  \centering
\begin{tikzpicture}
\tikzset{
  faceup/.style={rectangle, draw, thick, minimum width=1.2cm, minimum height=1.6cm, fill=white},
}

\newcommand{\stackdown}[3]{%
  \foreach \i in {1,...,#3}{%
    \fill[gray!40, draw=black, thick]
      (#1-0.6, {#2 - \i*0.5}) rectangle (#1+0.6, {#2 - \i*0.5 + 0.5});
  }%
}

%% ── Stock pile (top left, gray like face-down cards) ─────────────
\fill[gray!40, draw=black, thick] (-5.6, 2.6) rectangle (-4.4, 4.2);

%% ── Foundation: four dashed rectangles ───────────────────────────
\foreach \i in {0,1,2,3}{
  \draw[dashed, thick] (-0.4+\i*1.6, 2.6) rectangle (0.8+\i*1.6, 4.2);
}

%% ── Tableau columns, tops all at y=1.5 ───────────────────────────

% Column 1: 0 face-down, 1 face-up (heart)
\node[faceup] at (-5, {1.5 - 0.8}) {{\color{red}$\heartsuit$}};

% Column 2: 1 face-down, 1 face-up (heart)
\stackdown{-3}{1.5}{1}
\node[faceup] at (-3, {1.5 - 1*0.5 - 0.8}) {{\color{red}$\heartsuit$}};

% Column 3: A of clubs
\stackdown{-1}{1.5}{2}
\node[faceup, label={[font=\small]below:$A$}] at (-1, {1.5 - 2*0.5 - 0.8}) {$\clubsuit$};
\node[right] at (-1.3, {1.5 - 1 + 0.25}) {$X$};

% Column 4: 3 face-down, 1 face-up (heart)
\stackdown{1}{1.5}{3}
\node[faceup] at (1, {1.5 - 3*0.5 - 0.8}) {{\color{red}$\heartsuit$}};

% Column 5: 4 face-down, 1 face-up (heart)
\stackdown{3}{1.5}{4}
\node[faceup] at (3, {1.5 - 4*0.5 - 0.8}) {{\color{red}$\heartsuit$}};

% Column 6: B of clubs  
\stackdown{5}{1.5}{5}
\node[faceup, label={[font=\small]below:$B$}] at (5, {1.5 - 5*0.5 - 0.8}) {$\clubsuit$};

%% ── Curved arrows ─────────────────────────────────────────────────
% Arrow between col 1 and col 2
\draw[->, >=Stealth, thick]
 (-1.7, {1.5 - 2*0.5 - 1.6}) .. controls (-1.7, {1.5 - 2*0.5 - 2.2})
  and (-2.3, {1.5 - 2*0.5 - 2.2}) .. (-2.3, {1.5 - 1*0.5 - 1.6});

% Arrow between col 5 and col 6
\draw[->, >=Stealth, thick]
  (4.3, {1.5 - 5*0.5 - 1.6}) .. controls (4.3, {1.5 - 5*0.5 - 2.2})
  and (3.7, {1.5 - 5*0.5 - 2.2}) .. (3.7, {1.5 - 4*0.5 - 1.6});

\end{tikzpicture}
\caption{Example starting configuration of a solitaire game.} \label{fig:solitaire}
\end{figure}

In solitaire, it is
difficult to know whether a starting configuration of cards will lead to a win
just by looking at it. One way we can figure out if it is a winnable
configuration is by going through every single subsequent set of moves that
leads to a terminating position, and seeing if that final position is a win. For
instance in Figure \ref{fig:solitaire}, we can either start by moving card $A$
or card $B$.
Whichever choice we make first will lead to another set of moves for each of them,
and each of those moves will lead to more moves, until the game ends by either
winning or getting stuck. If we perform all of the possible moves after we move
card $A$ and all of them end in a stuck position, then we still don't know whether
moving card $B$ first would lead to a win. At this point, we've done a lot of
work for little
result. This is an example of how the brute force method is often too difficult
to carry out. However, if we set up our cards and then our friend said, ``I bet
if you move card $A$, and then card $X$, and then card....you will win," we can
follow their instructions to verify that sequence leads to a win, therefore
proving to us that the starting configuration of cards is winnable. In this
case, we only had
to move at most 52 cards to come to that conclusion, which means we performed a
constant number of actions to solve the solitaire problem (if we say this is
true for any deck of
$n$ cards, it would be a polynomial number of actions). However, unless we all
have a friend that can forsee the future, we cannot guarantee that we would try
the winnable solving sequence first. So, $NP$ is the class of problems where we can
confirm a given solution is correct in polynomial time, but we don't know of a
polynomial time algorithm to find that solution ourselves. 

Keep in mind that problems we know are in $P$ are also in $NP$, because we can
verify that their solutions (which we can find in polynomial time) are correct
in polynomial time. We will denote the problems that we cannot prove are or are
not in $P$ as $NP\setminus{P}$, which means the set difference of $NP$ and $P$, or
all problems in $NP$ that are not known to be in $P$. Generally, we consider the
class of problems in $NP\setminus{P}$ to be very difficult to solve, as it means
no one has ever discovered an algorithm that can find a correct solution in
polynomial time. The hardest $NP$ problems are known as \textit{NP-complete}
problems, where “complete” means the problem can be generalized to take the form
of every other problem in $NP$. So, if a polynomial time algorithm was discovered
for an \textit{NP-complete} problem, it would mean that every $NP$ problem could be solved
in polynomial time, and therefore all of those problems would actually just be
in $P$ \cite{Sipser_2006}.

\section{Graph Theory}

In discrete mathematics, a \textit{graph} is a collection of nodes called
\textit{vertices} connected by \textit{edges}. We denote $G = (V, E)$ where $G$ is
the graph, and it is composed of a set of vertices $V$ and a set of edges $E$.
Mathematicians like graphs because their structure makes them very effective for
modeling a variety of real-world situations such as social networks (where
vertices are individuals, and edges connect people to their acquaintances) or
even the spread of disease (where vertices are people and edges represent an
infection spreading from one person to another). Many graph problems involve
finding a \textit{path}, or a sequence
of edges that joins a sequence of distinct vertices. In Figure
\ref{fig:simple-graph}, there are 3 paths from vertex a to vertex c. 

\begin{figure}[H] \centering
  \begin{tikzpicture}[
      node distance=2cm,
      vertex/.style={circle, draw, thick, minimum size=8mm, inner sep=0pt}
    ]

    \node[vertex] (1) at (0, 2)   {a};
    \node[vertex] (2) at (2, 2.5) {b};
    \node[vertex] (3) at (2, 0)   {c};
    \node[vertex] (4) at (3.5, 1) {d};
    \node[vertex] (5) at (5.5, 3) {e};
    \node[vertex] (6) at (6.5, 0) {f};

    \draw[thick] (1) -- (2);
    \draw[thick] (1) -- (3);
    \draw[thick] (2) -- (4);
    \draw[thick] (3) -- (4);
    \draw[thick] (3) -- (6);
    \draw[thick] (4) -- (5);
    \draw[thick] (5) -- (6);

  \end{tikzpicture}
  \caption{Simple graph with 6
    vertices and 7 edges.} \label{fig:simple-graph}
\end{figure}

Graphs can be used to model more complicated problems by adjusting their
structure slightly. A basic graph problem has vertices and edges, but we can add
factors such as edge \textit{direction} going from one vertex to another, or
assigning \textit{weights} to the edges. A ``weight" in
this context is some value derived from the constraints of the problem. The weighted graph
can then be used to model problems where we want to maximize or minimize the weights.
For instance, we can construct a graph where each vertex represents a city and
each edge represents a highway from one city to another. If we were planning a
roadtrip, we could add a weight to each edge to signify the travel time of that
highway from one city to another. Then, if we were planning a road trip to all
the cities on the graph, we could find the fastest route by finding the path
with the smallest total weight. If Figure \ref{fig:weighted-graph} represents our road
trip, it is faster to take the edge with weight 4 from a $\rightarrow$ c than to take the
path a $\rightarrow$ b $\rightarrow$ d $\rightarrow$ c, which has a total
weight of 6. 

\begin{figure}[H] \centering
  \begin{tikzpicture}[
      vertex/.style={circle, draw, thick, minimum size=8mm, inner sep=0pt},
      weight/.style={midway, fill=white, inner sep=1pt, font=\small}
    ]

    \node[vertex] (1) at (0, 2)   {a};
    \node[vertex] (2) at (2, 2.5) {b};
    \node[vertex] (3) at (2, 0)   {c};
    \node[vertex] (4) at (3.5, 1) {d};
    \node[vertex] (5) at (5.5, 3) {e};
    \node[vertex] (6) at (6.5, 0) {f};

    \draw[->] (1) -- (2);
    \draw[->] (1) -- (3);
    \draw[->] (2) -- (4);
    \draw[->] (4) -- (3);
    \draw[->] (3) -- (6);
    \draw[->] (4) -- (5);
    \draw[->] (5) -- (6);

    \draw[thick] (1) -- node[weight, above]       {2} (2);
    \draw[thick] (1) -- node[weight, left]        {4} (3);
    \draw[thick] (2) -- node[weight, left]        {3} (4);
    \draw[thick] (3) -- node[weight, below]       {1} (4);
    \draw[thick] (3) -- node[weight, below]       {8} (6);
    \draw[thick] (4) -- node[weight, above left]  {4} (5);
    \draw[thick] (5) -- node[weight, right]       {5} (6);

  \end{tikzpicture}
  \caption{Directional graph from
    Figure \ref{fig:simple-graph} with weights.} \label{fig:weighted-graph}
\end{figure}

There are many scenarios in which we may also want to find the maximum total
weight of a graph. Many polynomial time algorithms exist for maximizing weights,
one of the more well known ones being the \textit{Hungarian} algorithm that runs
in $O(V^3)$ time \cite{hungarian}. Algorithms such as the Hungarian algorithm can
be helpful in many instances, such as assigning resources to entities. We can use
edges to draw connections between resources and entities, and assign weights
based on how desirable that resource is to each entity. This construction models
real-life instances such as assigning taxi drivers to waiting customers, or (as we will
find in this thesis) classes to students. Resource assignment problems are
actually extensively studied under another field of research as well called
\textit{mechanism design}.


\section{Mechanism Design}

Mechanism design has traditionally been studied by economists and game
theorists, though in recent years computer scientists have been joining
the field as well because mechanisms are often just algorithms. Mechanism design is the
construction of systems that “fairly” (under various definitions of the word)
address scenarios involving a host of participants with adverse or overlapping
goals. Some examples of problems that fall under this classification are as
follows: designing a voting system that successfully eliminates strategizing by
any party \cite{voting}, auctioning search engine advertisement spots in a way
that promotes honest bidding \cite{auctions}, assigning roommates from a pool of
students such that each student ends up with a roommate they prefer
\cite{irving}, and maximizing aid while allocating food donations to food banks
\cite{foodbank}. There are two subcategories of mechanism design relevant to this
thesis: stable matching and fair division.

\subsection{Stable Matching}

Many real world scenarios can be modelled as \textit{matching} problems, such as
assigning medical students to hospitals for residency \cite{nrmp}. The
standard (and heteronormative) matching problem is known as the Stable Marriage
Problem \cite{princeton-marriage}, in which the goal is to match an equal number
of men and women such that no participant remains uncoupled and everyone has
exactly 1 partner.  This is known as a
\textit{bipartite} matching problem, or one in which the market of participants
is divided into two disjoint sides and matched from one side to the other. Each
participant provides a list of preferences, ranking who on the other side they
would like to be matched with from most to least. This is known as
\textit{two-sided} preferences, where every member of both sides provides a
preference list. An essential aspect of the Stable Marriage
Problem is that the matching should prevent “cheating,” which occurs when a man
and a woman who weren’t matched to each other prefer to be together over their
assigned partners and will “elope,” leaving the other two participants alone.
This couple would be known as a \textit{blocking pair}, because their elopement
“blocks” the matching from being complete. Herein lies the stability aspect of
stable matching: we wish to find an algorithm that outputs a complete matching
of couples with no blocking pairs.

\begin{figure} \centering
  \begin{tikzpicture}[
      node/.style={circle, draw, minimum size=0.7cm, inner sep=0},
      box/.style={rectangle, draw, minimum width=2.5cm, minimum height=0.8cm, align=center}
    ]

    % Labels (raised higher above the nodes)
    \node at (1, 1.2) {\text{Men}};
    \node at (5, 1.2) {\text{Women}};

    % Left preference box (gap increased so dotted line is short)
    \node[box] at (-1.5, 0.5) {$1 > 6 > 5 > 2 \cdots$};

    % Right preference box
    \node[box] at (7.5, 0.5) {$1 > 4 > 2 > 3 \cdots$};

    % Men nodes
    \node[node] (m1) at (1,  0)   {1};
    \node[node] (m2) at (1, -1)   {2};
    \node[node] (m3) at (1, -2) {3};
    \node[node] (m4) at (1, -3) {4};
    \node[node] (m5) at (1, -4)   {5};
    \node[node] (m6) at (1, -5)   {6};

    % Women nodes
    \node[node] (w1) at (5,  0)   {1};
    \node[node] (w2) at (5, -1)   {2};
    \node[node] (w3) at (5, -2) {3};
    \node[node] (w4) at (5, -3) {4};
    \node[node] (w5) at (5, -4)   {5};
    \node[node] (w6) at (5, -5)   {6};

    % Dotted lines from boxes to node 1 (with explicit start/end to avoid overlap)
    \draw[dotted, thick] (0.25, 0.5) -- (0.65, 0);
    \draw[dotted, thick] (5.35, 0)  -- (5.75, 0.5);

    % Edges
    \draw (m1) -- (w1);
    \draw (m2) -- (w5);
    \draw (m3) -- (w2);
    \draw (m4) -- (w6);
    \draw (m5) -- (w4);
    \draw (m6) -- (w3);

    % Arrows (matching from diagram: crossing pattern)
    \draw[->] (m1) -- (w1);
    \draw[->] (m2) -- (w5);
    \draw[->] (m3) -- (w2);
    \draw[->] (m4) -- (w6);
    \draw[->] (m5) -- (w4);
    \draw[->] (m6) -- (w3);

  \end{tikzpicture}
  \caption{Bipartite
    directional graph illustrating the Stable Marriage problem.}
  \label{fig:stable-marriage}
\end{figure}

In 1962, David Gale and Lloyd Shapley published a seminal stable matching paper, in
which they proposed the \textbf{deferred acceptance} (DA) algorithm: a solution
that guarantees a stable matching for the Stable Marriage problem in polynomial
time \cite{gale_shapley}. Gale and Shapley also proved their DA algorithm could
solve the College Admissions problem: matching high school student applicants to
colleges. College Admissions differs slightly from Stable Marriage: it is a
\textit{many-to-one} matching in which colleges accept multiple students rather
than each college matching with a single student, and it has \textit{incomplete
  preferences} because students do not rank (apply to) every possible college that
exists. With a slight adaptation of the DA algorithm, it can solve this
version of College Admissions. Since 1962, the DA algorithm has been adapted to
fit many more problems with additional factors, including the residency problem.

Stable matching lays much of the groundwork for the work we do in this thesis.
In a sense, we want to match students to HUM sections in a way that is ``stable,"
though our understanding of stability will change as we continue to explore our
problem because students and HUM sections cannot ``cheat" on each other in the way
men and women can. Despite this, much of the research that our work builds upon is
based on altered versions of the DA algorithm. We will also develop a new notion
of stability that expands upon this traditional definition from Stable Marriage.

\subsection{Fair Division}

The difference between matching and fair division (also known as fair
allocation) is most often two-sided versus \textit{one-sided} preferences. One-sided
preferences means only one side of the bipartite market cares what they
are matched to. Because only one side has preferences, we no longer have the idea that
“cheating” could occur; instead, we evaluate fair division under \textit{envy}.
For instance, say we have a cake that is half chocolate and half vanilla, and we are dividing
the cake between people at a party. A person who prefers vanilla but receives
chocolate will be \textit{envious} of those who receive a vanilla slice. In a
fair division problem, the goal is to minimize the amount of envy between
participants and maximize the amount of \textit{welfare}, where welfare is the
benefit participants yield from the goods they receive (such as feeling
significantly more joy from vanilla over chocolate).

\begin{figure} \centering
  \begin{tikzpicture}[
      node/.style={circle, draw, minimum size=0.8cm, inner sep=0},
      ->, >=Stealth
    ]

    % Cake rectangle (white left portion)
    \draw[line width=1.5pt] (0, 0) rectangle (10, 3);

    % Shaded right portion
    \fill[gray!20] (5, 0) rectangle (10, 3);
    \draw[line width=1.5pt] (5, 0) rectangle (10, 3);

    % Orange vertical cut lines (no arrowheads, just lines)
    \draw[orange, thick, -] (0.75,   -0.3) -- (0.75,   3.8);
    \draw[orange, thick, -] (1.5, -0.3) -- (1.5, 3.8);
    \draw[orange, thick, -] (2.25, -0.3) -- (2.25, 3.8);
    \draw[orange, thick, -] (5, -0.3) -- (5, 3.8);
    \draw[orange, thick, -] (5.75, -0.3) -- (5.75, 3.8);

    % Black arrows from midpoints between cuts down to nodes
    % v1: left edge (0) to cut1 (1), midpoint = 0.5
    \draw[->] (0.5,  0) -- (0.5,  -2);
    % v2: cut1 (1) to cut2 (1.8), midpoint = 1.4
    \draw[->] (1,  0) -- (1.8,  -2);
    % c1: cut2 (1.8) to cut3 (2.6), midpoint = 2.2
    \draw[->] (1.75,  0) -- (3.1,  -2);
    % v3: cut4 (5.5) to cut5 (6.2), midpoint = 5.85
    \draw[->] (5.5, 0) -- (4.4,  -2);

    % Nodes row
    \node[node] (v1) at (0.5,  -2.7) {$v_1$};
    \node[node] (v2) at (1.8,  -2.7) {$v_2$};
    \node[node] (c1) at (3.1,  -2.7) {$c_1$};
    \node[node] (v3) at (4.4,  -2.7) {$v_3$};
    \node[node] (v4) at (5.7,  -2.7) {$v_4$};
    \node[node] (c2) at (7.0,  -2.7) {$c_2$};
    \node[node] (c3) at (8.3,  -2.7) {$c_3$};
    \node[node] (c4) at (9.6,  -2.7) {$c_4$};

    % Small arrows below v1, v2 (pointing down = happy)
    \node at (0.5, -3.5) {\smiley};
    \node at (1.8, -3.5) {\smiley};
    \node at (3.1, -3.5) {\frownie};
    \node at (4.4, -3.5) {\frownie};
  \end{tikzpicture}
  \caption{Cake cutting example where the left half of the cake is vanilla, the
  right half is chocolate, $v$ nodes represent vanilla people, and $c$ nodes 
represent chocolate people. In this case, $c_1$ and $v_3$ would envy each other.} \label{fig:cake-cutting}
\end{figure}

The cake-cutting example is the canon example for what is known as
\textit{divisible goods} \cite{genesis}, but there is also a lot of study into the allocation of
\textit{indivisible goods} where items cannot be cut up like a cake. The aforementioned
food bank example, allocating donated items (such as non-perishable cans of
food) to different food banks based on preferences they provide, is a fair
division problem with indivisible goods \cite{foodbank}.

In addition to envy and welfare, fair division will often be evaluated under
Pareto-optimality. An allocation is considered \textit{Pareto-optimal} if there
is no other allocation that is strictly better for at least one participant
while retaining the same welfare for the other participants \cite{pareto}. As more factors
are added to fair allocation problems—such as considering the order in which
goods arrive for distribution, scaling the number of participants, and defining more
specific versions of welfare—envy-freeness and Pareto-optimality become more
difficult to achieve.

Many problems like HUM section assignment are modelled as fair division
problems, as they involve one-sided preferences. This includes a basic version of
course allocation and an altered version of College Admissions, which we will
discuss in Section 1.5. We will use some of these concepts to define our problem
in Chapter 2, combining facets of stable matching and fair division.

\section{Adding Constraints}

What’s known as the \textit{vanilla} versions of these matching and division
environments, which is to say the more simple or basic environments without many
additional factors, have been
studied thoroughly by now. So, to advance mechanism design research and to more
accurately model real-world scenarios, scientists add \textit{constraints} to
the problems. Constraints come in many forms, and often make problems so complex
they become \textit{NP-complete}. In the scope of class assignment and related matching
problems, constraints often come in the forms of quotas on class capacity and
categories of students.

\subsection{Quotas and Diversity}

Quotas in allocation and matching problems can be implemented as upper and lower
bounds on the mechanism. Upper bound quotas imply capacity constraints, such as
a maximum number of open positions per hospital in the residency matching
problem. Gale and Shapley’s College Admissions problem, in which a college with
$n$ applicants can admit up to $q$ students, is traditionally studied with these
capacity constraints (where $q$ is for quota). Lower bound quotas in something
like the College Admissions would encode in the mechanism that schools need to
accept at least $m$ students \cite{quotas}. Strict quotas can cause
finding a stable matching in a bipartite setting
to be \textit{NP-hard} \cite{santhini}, so quotas are often studied under “soft”
constraints as well: for example, allowing a quota to be a target range rather
than a specific number.

The College Admissions problem can be extended to more accurately
reflect reality by modelling diversity constraints. The goal of diversity
constraints is usually to produce a matching/allocation in which some type of
participants is spread evenly or represented accurately in the result. In the
actual college admissions problem, these constraints are commonly known as
Affirmative Action and have traditionally been implemented in the form of
capacity quotas on majority students to give space for minority students. There
are many papers that mathematically
study that particular version of Affirmative Action, as well as prove that it does
not always benefit the minority applicants \cite{affirmative-action}. However,
diversity constraints can take many forms—they have been studied under
frameworks of lower-bound quotas \cite{quotas}, proportionality (evenly
distributing types of students in the matching) \cite{proportionality}, and
\textit{group fairness} (equalizing outcomes across protected and non-protected
groups) \cite{fairness}. In each of those papers implementing diversity
constraints, the constraints make the problems \textit{NP-complete}.

\subsection{Group Fairness}

``Group fairness" in classification or assignment tasks refers to
``preventing discrimination against against protected population subgroups"
while still achieving some ``utility" in the assignment \cite{fairness}. These problems
involve classifying individuals themselves by some metric of ``similarity" with
respect to the particular task. A lot of research has been done on defining
``similarity" mathematically, but in this thesis we are more concerned with
ensuring pre-determined similar students are treated fairly in class assignment.
Because the ``diversity" aspect of our HUM section assignment problem refers to academic
subject diversity, we will use ``group" to mean groups of majors (such as arts,
social sciences, natural sciences, etc.). That is, we use group fairness to
signify that students of different academic groups are represented proportionally within
HUM sections with respect to the overall population  of their group in the student
body. We can also think of group fairness in our problem as the constraint that students of
all academic groups at Reed have equal opportunity to be in all of the HUM sections.

\section{Related Work}

The problem we discuss in this thesis is related to two extensively-studied
problems: school choice and course allocation. School choice in mechanism design
is the problem of assigning high school students to high schools such that
parents are satisfied, each student has a school to attend, and no school admits
more students than they can accommodate. Course allocation in mechanism design
is the problem of constructing a system to assign students to classes based on
which classes they prefer to take.

\subsection{Two-sided matching}

In 2003, Abdulkadiroğlu and Sonmez published a foundational paper on school choice,
in which they view the problem through the bipartite lens of schools and
students each with distinct preferences.  This lens allows them to apply the
Gale-Shapley DA algorithm as a solution as well as their own algorithm, which
they called Top Trading Cycles \cite{school-choice}. In Top Trading
Cycles, if the matching has a cycle of students in which each student prefers
the next student’s school, each assignment is given to the next student until
the cycle is broken.

Six years later, in 2009, Abdulkadiroğlu, Pathak, and Roth amended the 2003
paper with a case study into the New York City high school allocation mechanism,
scrutinizing the DA algorithm under stability. They show
empirically that there is a loss of Pareto efficiency in the algorithm when
adding stability to the environment \cite{abu-2009}.

Many later papers extend one or
both of the aforementioned papers in further research. In 2010, Biró et al.
studied school choice in a college setting with both upper and lower quotas,
constraining the environment in a new way. They determined that stable matchings
do not always exist with those quotas, and in fact determining whether a stable
matching exists at all is \textit{NP-complete} \cite{quotas}. Three years later, Hafalir
et al. analyzed the DA algorithm for school choice under quotas for diversity,
where they found that using upper quotas produced a matching that was
Pareto-dominated by a matching with lower quotas for minority students
\cite{affirmative-action}. There has since been much more research into the
stability effects of quotas on the two-sided school choice problem, including
papers that implicate stability with diverse students of multiple types
\cite{kurata}.

\subsection{One-sided matching}

Since 2003, research also branched from two-sided school choice into one-sided
matching, where only one side of the market has preferences. This was often
studied under the framework of course allocation. In 2011, Budish and
Cantillon published the defining paper on course allocation at Harvard (which
cites Abdulkadiroğlu’s 2003 and 2009 papers) in which they evaluate Harvard’s
mechanism under Pareto-optimality \cite{budish}. There
have been slight alterations to their framework since then, such as the effects
of incomplete preferences on participant welfare.

Only in recent years have scholars started studying one-sided matching with
quotas and other diversity constraints. In fact, the fairness constraints closest to that
in this paper only
seem to exist in literature published as recently as March-August 2025. Specifically,
Santhini et al. researched the empirical effects of both soft quotas
\cite{santhini} and group fairness \cite{santhini-group-fairness} on the
mechanism using convex optimization (also done similarly by Panda et al.
\cite{pandagroupfairness}). There is not readily as much theory-based research
on one-sided matching with constraints as two-sided matching, and those that do
study it under a theoretical framework use more of a fair division envy/Pareto-based analysis than
stability.

As far as we can tell, the closest paper to our construction of stability in
this thesis is an unpublished 2011 paper by Abdulkadiroğlu, who generalizes
two-sided and one-sided preferences to a single environment in which the Top
Trading Cycles algorithm can be reduced to work for both kinds of preferences.
In the paper, he frames
stability to be “if, whenever a student prefers a school to her assignment, that
school is enrolled up to its capacity by students who have higher priority at
that school” \cite{abdu2011generalized}, where “higher priority” in the
one-sided setting only considers how strongly the other students prefer that
school. The paper does not consider the effect of adding fairness constraints.

