% !TEX root = ../thesis.tex
\chapter{Formalizing the Problem} 
The existing research
on school choice and course allocation would seem to suggest that for our
problem, which only features one-sided preferences as opposed to preferences on
both sides of the market, we follow a fair division approach similar to that of
Budish’s one-sided matching \cite{budish}. However, given the practical application of our
problem as a mechanism that is meant to reduce the Registrar's workload, we
determined that constructing a notion of stability makes more sense than
evaluating envy in a fair division manner. While we care about student happiness in our
solution, our formulation of the problem is primarily driven by the goal of
assigning year-long HUM sections that are balanced both in size and student
majors, meaning we want to minimize the ability of students to swap with each
other post-assignment so as not to disrupt the balance. If no two students
recipricolly wish to swap with each other, then we can consider the assignment stable. This is the
framework under which we shall formalize our model.

\section{The Model}

The basic components of our model are the two sides of our market: a set of
students and a set of HUM class sections, each of which have a certain number of
seats based on the class capacity. Though we are working in a bipartite setting,
only one side of our market has preference over the other (i.e., students prefer
certain classes, whereas classes are inanimate and therefore indifferent to
students...as far as we know). Students
also do not necessarily have ranked preference over every single available
section; they likely equally do not want to take any sections that conflict with
their other selected courses. So, preference lists will be incomplete, with
those sections that students have a conflict with not being included in their ranking.
\newpage

\begin{definition} \label{def:overview}
  The \emph{HUM-assignment} problem consists of a finite set of Reed College
  students, $\mathcal{R}$, and a finite set of class sections, $\mathcal{C}$.
  Each class section $C \in \mathcal{C}$ has a capacity $q$ representing the
  number of seats in the class: we can think of each $C$ as a set of seats where
  $|C| \leq q$. We denote the set of all seats $\mathcal{S}$, where $C \subset
  \mathcal{S}$ is disjoint for all $C \in \mathcal{C}$. Students rank classes in
  $\mathcal{C}$: each student $r \in \mathcal{R}$ has a strict preference relation $\succ_r$ over
  $\mathcal{C} \cup \{\emptyset\}$, where $\emptyset$ represents the outside
  option of a student. That is, $\emptyset \succ_r C$ indicates a student prefers
  no match over being matched with $C$ for all $C \in \mathcal{C}$.
\end{definition}

\begin{definition} \label{def:1m}
  A \emph{1-sided matching (1M)} of students to class sections is a function $\mu :
  \mathcal{R} \rightarrow \mathcal{C} \cup \emptyset $ such that
  \begin{itemize}
    \item
          Each student is assigned to a class section they prefer or no
          seat: \\ $\mu(r) \succ_r \emptyset\text{ or } \mu(r) = \emptyset
          \text{ for all }r \in \mathcal{R}$.
    \item
          No class exceeds its capacity:
          $|\{r \in \mathcal{R} : \mu(r) \in C\}| \leq q$ \text{ for all } $C
          \subset \mathcal{S}$.
  \end{itemize}
\end{definition}

Finding a 1-sided matching would solve an instance of the HUM-assignment problem
as defined above. However, we still have to introduce the last constraint of the
problem: \emph{group fairness}.

Students can be divided into three groups based on the way Reed College divides
subjects: Group 1 for language, arts, and literature, Group 2 for
social sciences, Group 3 for math and lab sciences \cite{group-reqs}. To implement the HUM
philosophy, which is that the class provides an opportunity for students from all
knowledge backgrounds and academic interests to discuss relevant texts, we want
to ensure a distribution of students that ``fairly" represents each group in every
HUM section. Practically, this means avoiding scenarios such as all Chem 101 students
with the same lab being placed
in the same section due to the overwhelming similarity in their schedules. This
will come in the form of \textit{quotas or caps} on the number of seats per section that
can be assigned to each group of student, meaning each class will be divided 
into two sets of seats. One set will be \textit{reserved} seats, representing
some number of seats in each class that are reserved for each group of student.
The other set is a number of \textit{free} seats in each class that any student can be
assigned to, which will allow for some flexibility. 

\begin{definition} \label{def:student-groups}
  Under the group fairness constraint, $\mathcal{R}$ is partitioned into
  three disjoint sets: $X, Y, Z \subset \mathcal{R}$, where $X \cup Y \cup Z
  \subseteq \mathcal{R}$
  and $X \cap Y \cap
    Z = \emptyset$. Denote students in $X$ as $x_1, x_2, ..., x_i$, students in
    $Y$ as $y_1, y_2, ..., y_j$, and students in $Z$ as $z_1, z_2, ..., z_k$.
\end{definition}

\begin{definition} \label{def:group-caps}
  For each class $C \in \mathcal{C}$ with capacity $q$, \[ |C| = q = \sum_{t \in
  \{X,Y,Z,F\}} q^t \] where $X \cup Y \cup Z \cup F \subseteq
C$. Similarly to Definition \ref{def:student-groups}, sets of seats $X, Y, Z$ correspond to
the three groups of students. We denote the set of all reserved seats \\
$\mathcal{S}_{res} = X \cup Y \cup Z$. $F$ represents the set of free seats. We
denote a specific seat $s \in C$ reserved for group $T$ as $s_c^T$. 
\end{definition}

\begin{definition} \label{def:f1m}
  A \emph{group-fair 1-sided matching (F1M)} builds on Definition \ref{def:1m}
  of a 1M by adding the constraints that no class is assigned more students of each
  group than there are reserved and free seats for those student where no more
  than one student is matched to each free seat, maintaining that no more
  total seats are matched for a class than its overall capacity:
  \begin{itemize}
    \item $|\{t \in T : \mu(t) \in C\}| \leq q^T + q^F$. 
    \item $|\{r \in \mathcal{R} : \mu(r) \in C\}| \leq q$ \text{ for all } $C \subset
      \mathcal{S}$.
    \item $|\{r \in \mathcal{R} : \mu(r) \in F \subset C\}| \leq q^F \text{ for all } C
      \subset \mathcal{S}$.
  \end{itemize}
\end{definition}

To make things clearer, we can visualize our problem using a graph (see Figure \ref{fig:model}).

\begin{figure}[H] \centering
\definecolor{plasmaA}{HTML}{0D0887}
\definecolor{plasmaB}{HTML}{CC4678}
\definecolor{plasmaC}{HTML}{F89441}
 \begin{tikzpicture}[
  classnode/.style={circle, draw, minimum size=1cm, inner sep=0},
  studentnode/.style={circle, draw, minimum size=0.8cm, inner sep=0, thick},
  ->, >=Stealth
]

%% ── Column headers ──────────────────────────────────────────────
\node at (-3, 1.8) {$\mathcal{C}$};
\node at (-3, 1.2) {\textsc{classes}};
\node at (-1.0, 1.8) {$\mathcal{S}$};
\node at (-1.0, 1.2) {\textsc{seats}};
\node at (3.5,  1.8) {$\mathcal{R}$};
\node at (3.5,  1.2) {\textsc{students}};

%% ── Preference box (top right) ──────────────────────────────────
\node[draw, minimum width=2.8cm, minimum height=0.7cm, align=center]
(prefbox) at (7.5, 0.2) {$C_1 \succ_{x_1} C_3 \succ_{x_1} C_n$};

%% ── Helper macro: horizontal coloured squares ───────────────────
\newcommand{\seats}[6]{%
  % #1=xcenter, #2=ytop, #3=nA(row1), #4=nB(row2), #5=nC(row3), #6=nF(row4)
  \tikzmath{
    \sqw = 0.35;
    \sqh = 0.35;
    \gap = 0.05;
    \hstep = \sqw + \gap;
    \vstep = \sqh + \gap;
    % Row 1: plasmaA (pink/red)
    for \j in {1,...,#3}{
      \xx = #1 + (\j-1)*\hstep;
      { \draw[plasmaA, thick] (\xx, #2) rectangle (\xx+\sqw, #2+\sqh); };
    };
    % Row 2: plasmaA (blue)
    for \j in {1,...,#4}{
      \xx = #1 + (\j-1)*\hstep;
      { \draw[plasmaB, thick] (\xx, #2-\vstep) rectangle (\xx+\sqw, #2-\vstep+\sqh); };
    };
    % Row 3: green
    for \j in {1,...,#5}{
      \xx = #1 + (\j-1)*\hstep;
      { \draw[plasmaC, thick] (\xx, #2-2*\vstep) rectangle (\xx+\sqw, #2-2*\vstep+\sqh); };
    };
    % Row 4: black
    for \j in {1,...,#6}{
      \xx = #1 + (\j-1)*\hstep;
      { \draw[black, thick] (\xx, #2-3*\vstep) rectangle (\xx+\sqw, #2-3*\vstep+\sqh); };
    };
  }
}
\newcommand{\seatsrow}[2]{\seats{#1}{#2}{2}{3}{4}{3}}

%% ── Class C1 ────────────────────────────────────────────────────
\node[classnode] (c1) at (-3, -0.2) {$C_1$};
\seatsrow{-2.0}{0.2}


%% ── Class C2 ────────────────────────────────────────────────────
\node[classnode] (c2) at (-3, -2.6) {$C_2$};
\seatsrow{-2.0}{-2.2}

%% ── Vertical dots ────────────────────────────────────────────────
\node at (-3, -4.2) {$\vdots$};

%% ── Class Cn-1 ──────────────────────────────────────────────────
\node[classnode] (cn1) at (-3, -6) {$C_{n-1}$};
\seatsrow{-2.0}{-5.6}

%% ── Class Cn ────────────────────────────────────────────────────
\node[classnode] (cn) at (-3, -8.4) {$C_n$};
\seatsrow{-2.0}{-8}

%% ── Student nodes: group X ───────────────────────────────────────
\node[studentnode, plasmaA, text=plasmaA] (a1) at (3.5,  0.45) {$x_1$};
\node[studentnode, plasmaA, text=plasmaA] (a2) at (3.5, -0.45) {$x_2$};
\node at (3.5, -1.1) {$\vdots$};
\node[studentnode, plasmaA, text=plasmaA] (ai) at (3.5, -2) {$x_i$};

%% ── Student nodes: group Y ───────────────────────────────────────
\node[studentnode, plasmaB, text=plasmaB] (b1) at (3.5, -3.1) {$y_1$};
\node[studentnode, plasmaB, text=plasmaB] (b2) at (3.5, -4) {$y_2$};
\node at (3.5, -4.65) {$\vdots$};
\node[studentnode, plasmaB, text=plasmaB] (bj) at (3.5, -5.65) {$y_j$};

%% ── Student nodes: group Z ───────────────────────────────────────
\node[studentnode, plasmaC, text=plasmaC] (cc1) at (3.5, -6.85) {$z_1$};
\node[studentnode, plasmaC, text=plasmaC] (cc2) at (3.5, -7.75) {$z_2$};
\node at (3.5, -8.45) {$\vdots$};
\node[studentnode, plasmaC, text=plasmaC] (ck) at (3.5, -9.35) {$z_k$};

%% ── Curly braces for groups ──────────────────────────────────────
\draw[-, decorate, decoration={brace, amplitude=6pt}, plasmaA]
  (4.2, 0.85) -- (4.2, -2.5) node[midway, right=8pt, plasmaA] {$X$};
\draw[-, decorate, decoration={brace, amplitude=6pt}, plasmaB]
  (4.2, -2.8) -- (4.2, -6.15) node[midway, right=8pt, plasmaB] {$Y$};
\draw[-, decorate, decoration={brace, amplitude=6pt}, plasmaC]
  (4.2, -6.45) -- (4.2, -9.8) node[midway, right=8pt, plasmaC] {$Z$};

%% ── Arrow from a1 to preference box ─────────────────────────────
\draw[-, dotted] (a1.east) -- (prefbox.west);
\node[above, yshift=10pt, xshift=15pt, font=\small] at ($(prefbox.west)$) {$P_r(x_1)$};

%% ── Crossing edges (seats -> students) ──────────────────────────
\draw[-] (-1.85,  0.4) -- (a1);
\draw[-] (-1.45,  0.35) -- (a2);
\draw[-] (-1.85,  0.0) -- (b1);
\draw[-] (-1.85, -2.4) -- (b2);
\draw[-] (-1.85, -5.45) -- (ai);
\draw[-] (-1.85, -6.20) -- (cc1);
\draw[-] (-1.85, -8.65) -- (cc2);
\draw[-] (-1.85, -8.25) -- (bj);
\draw[-] (-1.45, -8.65) -- (ck);

\end{tikzpicture}\caption{Graph illustrating an F1M, where each group of
students, $X, Y,$ and $Z$, is represented by a different color.} \label{fig:model}
\end{figure}
In a graph representing the HUM-assignment problem, the vertices are comprised
of students and seats that belong to specific class sections, and an edge is
drawn between a student and a seat if
the corresponding class is on the student's preference list. We can formally define this
representation.

\begin{definition}
  Let $G = (V, E)$ be a bipartite graph where:
  \begin{itemize}
    \item
      $V = \mathcal{R} \cup \mathcal{S}$
    \item
      $E = \{(t, s_c^T) \mid t \in T \subset \mathcal{R}, s_c^T \in \mathcal{S}, C \in
        \mathcal{C}, C \succ_t
            \emptyset\}$.
  \end{itemize}
\end{definition}

In this form, the graph only represents our problem, not a solution. For a graph
to represent a solution or a matching, each student would only have one edge
connecting them to a single seat in a class in a way that adheres to class capacities. So,
we need a way to narrow down which edges lead to a viable solution. That means
we need a way to judge edges against each other: we need to assign value to
them. To accomplish this, we can give each edge a weight that represents how
conducive it is to the constraints of our problem based on how well that match
would accomplish our goals. Recall that we want to satisfy two goals: every
student is assigned to a class they prefer, and all reserved seats are filled by
students of that reserved group. We
call a matching \textit{full} if every student is matched to a seat they prefer. 
Because these two goals are equally important to us, we can simply increment
the weight of each edge by 1 for each goal it satisfies. We can now supplement
our previous definition of a graph.

\begin{figure}[H] \centering
\definecolor{plasmaA}{HTML}{0D0887}
\definecolor{plasmaB}{HTML}{CC4678}
\definecolor{plasmaC}{HTML}{F89441}
  \begin{tikzpicture}[
  classnode/.style={circle, draw, minimum size=1cm, inner sep=0},
  studentnode/.style={circle, draw, minimum size=0.8cm, inner sep=0, thick},
  weight/.style={midway, fill=white, inner sep=1pt, font=\small},
  ->, >=Stealth
]

%% ── Column headers ──────────────────────────────────────────────
\node at (-3, 1.8) {$\mathcal{C}$};
\node at (-3, 1.2) {\textsc{classes}};
\node at (-1.0, 1.8) {$\mathcal{S}$};
\node at (-1.0, 1.2) {\textsc{seats}};
\node at (3.5,  1.8) {$\mathcal{R}$};
\node at (3.5,  1.2) {\textsc{students}};

%% ── Preference box (top right) ──────────────────────────────────
\node[draw, minimum width=2.8cm, minimum height=0.7cm, align=center]
(prefbox) at (7.5, 0.2) {$C_1 \succ_{x_1} C_2 \succ_{x_1} C_n$};

%% ── Helper macro: horizontal coloured squares ───────────────────
\newcommand{\seats}[6]{%
  % #1=xcenter, #2=ytop, #3=nA(row1), #4=nB(row2), #5=nC(row3), #6=nF(row4)
  \tikzmath{
    \sqw = 0.35;
    \sqh = 0.35;
    \gap = 0.05;
    \hstep = \sqw + \gap;
    \vstep = \sqh + \gap;
    % Row 1: plasmaA (pink/red)
    for \j in {1,...,#3}{
      \xx = #1 + (\j-1)*\hstep;
      { \draw[plasmaA, thick] (\xx, #2) rectangle (\xx+\sqw, #2+\sqh); };
    };
    % Row 2: plasmaA (blue)
    for \j in {1,...,#4}{
      \xx = #1 + (\j-1)*\hstep;
      { \draw[plasmaB, thick] (\xx, #2-\vstep) rectangle (\xx+\sqw, #2-\vstep+\sqh); };
    };
    % Row 3: green
    for \j in {1,...,#5}{
      \xx = #1 + (\j-1)*\hstep;
      { \draw[plasmaC, thick] (\xx, #2-2*\vstep) rectangle (\xx+\sqw, #2-2*\vstep+\sqh); };
    };
    % Row 4: black
    for \j in {1,...,#6}{
      \xx = #1 + (\j-1)*\hstep;
      { \draw[black, thick] (\xx, #2-3*\vstep) rectangle (\xx+\sqw, #2-3*\vstep+\sqh); };
    };
  }
}
\newcommand{\seatsrow}[2]{\seats{#1}{#2}{2}{3}{4}{3}}

%% ── Class C1 ────────────────────────────────────────────────────
\node[classnode] (c1) at (-3, -0.2) {$C_1$};
\seatsrow{-2.0}{0.2}


%% ── Class C2 ────────────────────────────────────────────────────
\node[classnode] (c2) at (-3, -2.6) {$C_2$};
\seatsrow{-2.0}{-2.2}

%% ── Vertical dots ────────────────────────────────────────────────
\node at (-3, -4.2) {$\vdots$};

%% ── Class Cn-1 ──────────────────────────────────────────────────
\node[classnode] (cn1) at (-3, -6) {$C_{n-1}$};
\seatsrow{-2.0}{-5.6}

%% ── Class Cn ────────────────────────────────────────────────────
\node[classnode] (cn) at (-3, -8.4) {$C_n$};
\seatsrow{-2.0}{-8}

%% ── Student nodes: group X ───────────────────────────────────────
\node[studentnode, plasmaA, text=plasmaA] (a1) at (3.5,  0.45) {$x_1$};
\node[studentnode, plasmaA, text=plasmaA] (a2) at (3.5, -0.45) {$x_2$};
\node at (3.5, -1.1) {$\vdots$};
\node[studentnode, plasmaA, text=plasmaA] (ai) at (3.5, -2) {$x_i$};

%% ── Student nodes: group Y ───────────────────────────────────────
\node[studentnode, plasmaB, text=plasmaB] (b1) at (3.5, -3.1) {$y_1$};
\node[studentnode, plasmaB, text=plasmaB] (b2) at (3.5, -4) {$y_2$};
\node at (3.5, -4.65) {$\vdots$};
\node[studentnode, plasmaB, text=plasmaB] (bj) at (3.5, -5.65) {$y_j$};

%% ── Student nodes: group Z ───────────────────────────────────────
\node[studentnode, plasmaC, text=plasmaC] (cc1) at (3.5, -6.85) {$z_1$};
\node[studentnode, plasmaC, text=plasmaC] (cc2) at (3.5, -7.75) {$z_2$};
\node at (3.5, -8.45) {$\vdots$};
\node[studentnode, plasmaC, text=plasmaC] (ck) at (3.5, -9.35) {$z_k$};

%% ── Curly braces for groups ──────────────────────────────────────
\draw[-, decorate, decoration={brace, amplitude=6pt}, plasmaA]
  (4.2, 0.85) -- (4.2, -2.5) node[midway, right=8pt, plasmaA] {$X$};
\draw[-, decorate, decoration={brace, amplitude=6pt}, plasmaB]
  (4.2, -2.8) -- (4.2, -6.15) node[midway, right=8pt, plasmaB] {$Y$};
\draw[-, decorate, decoration={brace, amplitude=6pt}, plasmaC]
  (4.2, -6.45) -- (4.2, -9.8) node[midway, right=8pt, plasmaC] {$Z$};

%% ── Arrow from a1 to preference box ─────────────────────────────
\draw[-, dotted] (a1.east) -- (prefbox.west);
\node[above, yshift=10pt, xshift=15pt, font=\small] at ($(prefbox.west)$) {$P_r(x_1)$};

%% ── Crossing edges (seats -> students) ──────────────────────────
% From C1 seats area to students
\draw[-] (-1.85,  0.4) -- node[weight, above] {2}(a1);
\draw[-] (-1.85,  -0.85) -- node[weight, above] {1}(a1);
% From C2
\draw[-] (-1.85, -2.05) -- node[weight, above] {2}(a1);
\draw[-] (-1.85,  -3.25) -- node[weight, below right] {1}(a1);
% From Cn
\draw[-] (-1.85, -7.85) -- node[weight, above left] {2}(a1);
\draw[-] (-1.85, -9.05) -- node[weight, below right] {1}(a1);

\end{tikzpicture}
\caption{Graph illustrating weighted edges for student $x_1$.} \label{fig:weighted-example}
\end{figure}

\begin{definition}
  \label{weighted-def}
  Let $G = (V, E)$ be a bipartite graph where:
  \begin{itemize}
    \item
           $V = \mathcal{R} \cup \mathcal{S}$
    \item
          $E = \{(t, s_c^T) \mid t \in \subset \mathcal{R}, s_c^T \in \mathcal{S}, C \in
        \mathcal{C}, C \succ_t
            \emptyset\}$
    \item
          $w : E \rightarrow \{1 , 2 \} $ is a weight function
          defined as \[ w(t, s_c^T) = \mathbf{1}\{\text{if student $t$ prefers class
          $C$}\} + \mathbf{1}\{\text{$s_c^T$ is reserved for $t \in T \subset
        \mathcal{R}$}\}. \]
  \end{itemize}
\end{definition}

Now, we have a way to compare edges to each other to allow us to find a
solution if one exists. We start with multiple edges between each student and
various reserved and free seats in classes on their list, as shown in Figure \ref{fig:weighted-example}. However, we want only
one edge between each student and a seat in the class section they will be
assigned to in the final matching. Based on the way we've set up our weights, we know that
the ``best" available solution (which satisfies the most constraints) will be the
graph with the maximum total weight when
only one edge remains for each student. In Figure
\ref{fig:weighted-example}, we can observe how maximizing the total weight works
in practice: if we want to preserve just one edge for student $x_1$ when
matching them to $C_1$, we would keep the edge weighted 2 to the reserved seat
rather than the weight 1 edge to the free seat. Our maximized total weight matching then has the
properties required by our F1M (Definition \ref{def:f1m}), being that students of each group
are prioritized assignments to reserved seats in classes they prefer, ensuring each
group is represented fairly within each class.

\begin{definition}
  A matching $\mu$ is a \emph{maximum-weight matching}
  if for all F1M matchings $\mu'$: \[ \sum_{r \in R} w(r, \mu(r)) \geq \sum_{r
      \in R} w(r, \mu'(r)) \]
\end{definition}

Observe that a maximum-weight matching would not only serve as a F1M
solution, but it would also (by nature of the way we assign weights based on
student preferences) ensure the solution is one in which the greatest number
of students are assigned to a class on their preference list. So, viewing the
problem through the lens of a weighted graph provides us with a helpful
framework for satisfying the problem.

\section{The Challenge of Stability} 
We've determined how to find a solution to the F1M problem from Definition
\ref{def:f1m}, but there is one more aspect to the problem we must consider.
Recall that matching is often studied under the analysis of
\textit{stability}. Notably, “stability” in our case adopts a different
meaning than in a traditional stable matching problem. Because we are
working with one-sided preferences, we do not consider “cheating” as a
violation of stability between students and classes, but rather between
students themselves. That is, we can consider a matching stable if there are
no blocking pairs \emph{between students}.

\begin{definition}
  \label{def:blocking-pair}
  A matching $\mu$ contains a  \textit{blocking pair} between
  students $r_1, r_2 \in \mathcal{R}$ if $\mu(r_2) \succ_{r_1} \mu(r_1)$ and
  $\mu(r_1) \succ_{r_2}
    \mu(r_2)$.
\end{definition}


\begin{definition}
  \label{def:stability}
  A matching $\mu$ is \emph{stable} if there exists no blocking pair between any
  two students $r_1, r_2 \in \mathcal{R}$.
\end{definition}

This notion of stability proves difficult for our problem. If we look
back at our original definition of the HUM-assignment problem (Definition
\ref{def:overview}) with class capacities and
incomplete preferences, we will find that in some instances it is possible there
may not even exist a stable matching to find.

\begin{theorem} \label{thm:unstable}
  There is not always a stable matching with quotas and
  incomplete preferences.
\end{theorem}

\begin{proof}
  We prove by counterexample that a stable matching may not exist under quota
  constraints and incomplete preference lists. 
  Let $\mathcal{C} = \{ C_1, C_2\}$, where there are two groups of student
  $X$ and $Y$ such that $q^X, q^Y = 1$ for each $C \in \mathcal{C}$ (each class can
  have at most one student of
  each group). Let $\mathcal{R} = \{ x_1, y_1, x_2, y_2 \}$. Let the preference lists be as
  follows: \[
    \begin{array}{ll} x_1: & C_1           \\ y_1: & C_2 \\ x_2: & C_1 \succ C_2 \\
             x_2:      & C_2 \succ C_1
    \end{array}
  \]

  In this case, regardless of the order of matching or algorithm, there is no
  stable matching. The $X$ students will both wish to be matched to $C_1$, which
  only has one $X$ seat, and both $Y$ students wish to be matched to $C_2$
  which only has one $Y$ seat. So, that means that in any matching, there is
  still an open $Y$ seat in $C_1$ and an open
  $X$ seat in $C_2$, which the unhappy $X$ and $Y$ students can then switch
  into on their own, respectively.
\end{proof}

Defining stability in this way is, then, too strict. Recall that even
determining whether there is a stable matching or not is \textit{NP-complete} in many
scenarios similar to this one \cite{santhini}. So, we need to relax our constraints
to allow us to explore some less restrictive notion of stability. For now, let’s
simplify the problem. At a basic level,
before we even think about stability, we want to work in an environment where we
can be sure an algorithm would respect all student preferences, such that no student
is assigned to a class they did not list as an option. 

To temporarily remove our notion of stability, we can alter our preferences from a
ranked system to what is known as \textit{0/1 preferences}. In 0/1 preferences,
students give a class a 1 if they are okay with being matched to it and a 0 if
it is not an option for them. This way,
there cannot be a notion of stability because all preferred classes have the
same rank. Notably, this simplified environment still lends itself well to the same
weighted graph construction from Definition \ref{weighted-def}, so we can
continue to use a weighted graph to represent our problem.

\begin{theorem}
  \label{max-weight} A matching $\mu$ fills all reserved seats while only
  assigning students to classes they prefer if and only if $w(\mu) =
  |\mathcal{R}| + |\mathcal{S}_{res}|$.
\end{theorem}

\begin{proof}
  First we will prove the forward direction. Let $\mu$ be a matching that
  assigns every student to a seat they
  prefer while also filling reserved seats first, such that all reserved seats
  are filled before free seats. By construction of the weights from Definition
  \ref{weighted-def}, we can determine a lower and upper bound for $w(\mu)$. 

  \begin{enumerate}
    \item
          \textbf{Lower bound of $\mu$:} Each edge exists only
          if the student prefers that seat, so every edge in $\mu$ contributes at
          least 1. By our assumption of characteristics of $\mu$, every student in
          $\mu$ got a seat they prefer, so summing over all students contributes at
          least $|\mathcal{R}|$ to $w(\mu)$. We also assumed each reserved seat was assigned to
          a student in $\mu$, which increments the weight by $+1$ for every
          edge from each of those reserved seats. Summing over all reserved seats gives an
          additional contribution of $|\mathcal{S}_{res}|$, because the rest of the seats are still
          1-weight edges. Therefore, we can bound $\mu$: \[ w(\mu) \geq
            |\mathcal{R}| +
            |\mathcal{S}_{res}|.
          \]

    \item
      \textbf{Upper bound of $\mu$:} We know the maximum edge
          weight between each student and seat is at most 2. Similarly to the
          lower bound, we know $\mu$ has at most $|\mathcal{R}|$
          edges because each student was matched to one or none of their preferences.
          However, because only the reserved seats contribute an additional weight
          to an edge, we know that the number of 2-weight edges $\leq
          |\mathcal{S}_{res}|$. So,
          the maximum summed weight of all edges must be $2|\mathcal{S}_{res}| +
          |\mathcal{R}| -
          |\mathcal{S}_{res}|$,
          representing the maximum number of 2-weight edges plus all students
          assigned to an overflow seat. Note that is equivalent to $|\mathcal{R}| +
          |\mathcal{S}_{res}|$. 
          Therefore, we can bound $\mu$: \[ w(\mu) \leq |\mathcal{R}| +
          |\mathcal{S}_{res}|. \]
  \end{enumerate}

  Note that any possible matching must then have a total weight $w \leq |\mathcal{R}| +
  |\mathcal{S}_{res}|$. Thus, if $w(\mu) = |\mathcal{R}| +
  |\mathcal{S}_{res}|$, $\mu$ is a maximum-weight matching with a lower bound of $|\mathcal{R}| +
  |\mathcal{S}_{res}|$ and an upper bound of $|\mathcal{R}| +
  |\mathcal{S}_{res}|$. Therefore, it must be that a matching that fully
  respects student preferences and seat reservations has a weight $w(\mu) =
  |\mathcal{R}| + |\mathcal{S}_{res}|$.

  Now, we prove the backwards direction. Let $\mu$ be a matching
  where $w(\mu) =  |\mathcal{R}| + |\mathcal{S}_{res}|$. Observe that $|E| \leq
  \mathcal{R}$, as $\mathcal{R}$ makes up one side of the market, so there can
  be at most $\mathcal{R}$ matches. The number of 2-weight edges then must be
  greater than or equal to 
  $|\mathcal{S}_{res}|$, because $w(\mu) - |\mathcal{R}| \leq |\mathcal{R}| +
  |\mathcal{S}_{res}| -  |\mathcal{R}| =  |\mathcal{S}_{res}|$. However, we know
  that there are at most $|\mathcal{S}_{res}|$ possible 2-weight edges, so in
  fact if the number of seats matched is greater than and less than
  $|\mathcal{S}_{res}|$, then we know exactly $|\mathcal{S}_{res}|$ seats were
  matched. Thus, a maximum-weight matching where $w(\mu) = |\mathcal{R}| +
  |\mathcal{S}_{res}|$ is a
  matching that fills all reserved seats while only assigning students to
  classes they prefer.
\end{proof}

Theorem \ref{max-weight} shows us that we can identify a way to 
assign each student to a class they prefer while respecting group fairness
constraints if such a matching exists. From here, we can begin to reintroduce
stability. While it may not be feasible to find stability between every student,
perhaps we can start by comparing just students of the same group.  We are
introducing a new notion of stability here, where we will consider a matching
stable if it exhibits \textit{intra-group stability}. This
means that for a student of any group, that student does not form a blocking pair
with another student \textit{of the same group}.

\begin{definition} \label{def:intra}
  A matching $\mu$ is \emph{intra-group stable} if, for any two students $t_1$
  and $t_2$ for all
  groups $T \subset \mathcal{R}$, $\mu(t_1) \succ_{t_1} \mu(t_2)$  or $\mu(t_2)
  \succ_{t_2}
    \mu(t_1)$. In other words, there exist no intra-group blocking pairs.
\end{definition}

With our new goal of intra-group stability, we can now add to our
understanding of a viable solution to our HUM-assignment problem.

\begin{definition} \label{def:ispim}
  An \emph{intra-group stable, group-fair 1-sided matching (ISF1M)} is a function $\mu :
  \mathcal{R} \rightarrow \mathcal{C}$ such that
  \begin{itemize}
    \item
          Each student is assigned to a class they prefer (the matching is full): \\
          $\mu(r) \succ_r \emptyset \text{ for all }r \in \mathcal{R}$.
    \item
          No class exceeds its capacity:  
          $|\{r \in \mathcal{R} : \mu(r) \in C\}| \leq q$ \text{ for all } $C \subset
      \mathcal{S}$.
    \item
          No class exceeds its capacity for each student group and free seats: \\ 
           $|\{t \in T : \mu(t) \in C\}| \leq q^T + q^F$ and
           $|\{r \in \mathcal{R} : \mu(r) \in F \subset C\}| \leq q^F \\ \text{ for all } C
      \subset \mathcal{S}$.
    \item
          There exist no intra-group blocking pairs: \\ for all $t_1, t_2 \in T
          \subset \mathcal{R}$, 
          $\mu(t_1) \succ_{t_1} \mu(t_2)$ or $\mu(t_2) \succ_{t_2} \mu(t_1)$
  \end{itemize}
\end{definition}

For now, our goal will be to develop an algorithm that finds an ISF1M to solve
our problem. We will classify our previous definition of stability (Definition
\ref{def:stability}) as \textit{inter-group stability} (blocking pairs involving
students of different groups) to draw a distinction between the two notions.
For now, we will consider an ISF1M to be a successful solution to the
HUM-assignment problem even though it is not inter-group stable. Though we would
ideally guarantee stability between all students, looking at the
problem through the lens of intra-group stability allows us to give some
theoretical guarantee about minimizing the amount of swapping that will occur
in an assignment while still remaining computationally feasible, as opposed to turning
to less-reliable heuristic solutions in an attempt to solve the broader conception of the problem.

